---
title: "merge_datasets"
author: "Chenyue Lu"
date: "4/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, include = FALSE}}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(dplyr)
```


```{r Read Data, include=FALSE}
## Covid Tracking Project Data
covid_tracking_url_path <- 'https://covidtracking.com/api/v1/states/daily.csv'
download.file(covid_tracking_url_path, destfile='covid_tracking_history.csv')
dat <- read_csv("../data/covid_tracking_history.csv")
population <- read_csv("../data/StatePopulation.csv")

# convert date into factor
dat$date <- as.factor(dat$date)

## School closure data
school_closure_data <- read_csv(url("https://raw.githubusercontent.com/luchenyue95/HMS_2020_BMI706_group6/master/data/coronavirus-school-closures-data-4-7-2020.csv"))
school_closure_data$StateClosureStartDate <- as.Date(school_closure_data$StateClosureStartDate, format = '%m/%d/%Y')

# Political makeup data
politics <- read.csv("../data/KFF_politics.csv",skip = 2, header = T)
# remove Footnotes column
politics <- politics[,!(colnames(politics) =='Footnotes')]
# remove more footnotes
politics <- politics[2:52,]
# convert full state names to abbreviations
politics$state_abb <- state.abb[match(politics$Location,state.name)]
population$state_abb <- state.abb[match(population$NAME, state.name)]
# state.abb doesn't have DC or PR. manually add this
politics[politics$Location=='District of Columbia',]$state_abb <- 'DC'
population[population$NAME == 'District of Columbia',]$state_abb <- 'DC'
population[population$NAME == 'Puerto Rico Commonwealth',]$state_abb <- 'PR'
```

#### Combining data

By combining the two data sets, one could derive the following information:

- How timing of school closures affected the rate of new cases overall and by state  
- Comparing 'R0' or the positive case rate before and after an intervention policy like school closures is enacted
- The 'incubation time' between an intervention policy like school closures and when there is an effect in  the number of cases or hospitalizations  
- How states' testing capacity correlated with school closure policies

Our goal is to explore whether school closures demonstrate direct impact on the rate of new cases.

```{r}
## remove NAs in school closure data - removed Bureau of Indian Education 
school_closure_data <- na.omit(school_closure_data, cols="StateClosureStartDate")

# add ntiles
n <- 3
school_closure_data$ClosureDateCat <- ntile(school_closure_data$StateClosureStartDate, n)
school_closure_data$ClosureDateCat <- factor(school_closure_data$ClosureDateCat, labels = c("Early", "Middle", "Late"), ordered = T)

## match up states - removed Department of Defense Education Activity
school_closure_data <- school_closure_data[school_closure_data$StateAbbreviation %in% unique(dat$state), ]

# reformat dates for merging
dat$date <- as.Date(dat$date,
  format = "%Y%m%d")
# merge two datasets by state
dat <- merge(dat,school_closure_data[c('StateAbbreviation','StateClosureStartDate')], by.x="state", by.y="StateAbbreviation")

# add into political data
dat <- merge(dat, politics, by.x='state','state_abb', all.x = TRUE)

# add column to indicate whether state was closed on that day 
dat <- dat %>%
  mutate(closed = date >= StateClosureStartDate) %>% # add column "closed" to dat based on whether it is closed that day
  arrange(desc(date))

# add region data
dat <- merge(dat, data.frame(state = state.abb, Region = state.region), by = 'state')

# add population data
dat <- merge(dat, population %>% select(state_abb, POPESTIMATE2019), by.x = 'state', 'state_abb', all.x = TRUE)
```

```{r Save}
write_csv(dat, "../data/merged_data.csv")
```
